---
title: "EDA Workflow — Interactive Demo"
subtitle: "DSPG 2026 | Virginia Tech"
author: "Yuanyuan Wen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: arrow
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5)
library(tidyverse)
library(palmerpenguins)
library(nycflights13)
library(skimr)
library(GGally)
library(corrplot)
library(scales)
library(patchwork)

state_data <- read_csv("../../data/state_acs_2022.csv", show_col_types = FALSE)
```

---

## Step 1 — Summary Statistics

### `skimr::skim()` — One Function, Full Picture

```{r skim-demo}
penguins |> skim()
```

### But `skim()` Is Not Enough — Anscombe's Quartet

Four datasets with nearly identical summary statistics:

```{r anscombe-table}
anscombe_long <- anscombe |>
  pivot_longer(everything(), names_to = c(".value", "dataset"),
               names_pattern = "(.)(.)") |>
  mutate(dataset = paste0("Dataset ", dataset))

anscombe_long |>
  group_by(dataset) |>
  summarise(
    `mean(x)` = round(mean(x), 2),
    `mean(y)` = round(mean(y), 2),
    `sd(x)`   = round(sd(x), 2),
    `sd(y)`   = round(sd(y), 2),
    `r(x, y)` = round(cor(x, y), 3)
  ) |>
  knitr::kable()
```

### The Reveal — Always Plot

```{r anscombe-plot}
ggplot(anscombe_long, aes(x = x, y = y)) +
  geom_point(color = "#861F41", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "#E5751F", linewidth = 1) +
  facet_wrap(~ dataset, nrow = 1) +
  theme_minimal(base_size = 13) +
  labs(subtitle = "Same statistics. Completely different data.")
```

---

## Step 2 — Spot Problems in the Numbers

### High Missingness and Impossible Values

```{r flag-problems}
# 1. Which variables have meaningful missingness?
penguins |>
  summarise(across(everything(), ~ mean(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  arrange(desc(pct_missing))

# 2. Check value ranges — should never be negative or > 100
state_data |>
  filter(pct_college < 0 | pct_college > 100) |>
  select(NAME, pct_college)

# 3. Check for impossible values specific to your data
penguins |>
  filter(body_mass_g < 0 | bill_length_mm > 200)
```

### Logical Consistency Checks

```{r logical-check}
# Pattern: components should sum to the total
# Example — percentage columns that don't sum to 100%
# across categories, end_date before start_date, etc.

# One sentence rule: for every derived variable, write the formula
# and verify it holds for all rows.

# Example with ACS data (if you have male + female breakdowns):
# acs_wide |>
#   mutate(
#     pct_recalculated = (col_male + col_female) / pop_total * 100,
#     discrepancy      = abs(pct_recalculated - pct_college)
#   ) |>
#   filter(discrepancy > 1)
```

---

## Step 3 — Hidden Missing Values

### Explicit vs Implicit Missing

```{r explicit-implicit}
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

# Q4 2020: explicitly missing (NA present)
# Q1 2021: implicitly missing (row never appears)
stocks |> complete(year, qtr)
```

### Sentinel / Placeholder Codes

```{r sentinel-codes}
# Flag and replace sentinel values before analysis
# Common codes: 0, 6, 9, 99, -99, -999, 9999

# df |> mutate(
#   income = na_if(income, 9999),
#   age    = na_if(age, 99)
# )

# Example: any zero incomes in state_data?
state_data |> filter(median_income == 0)
```

### Logically Missing Values

Some absences reflect real-world structure, not data errors:

- Wage data is missing for unemployed individuals — *because* they have no wage, not at random
- **Confidential suppression** — USDA agricultural reports omit data when too few farms report,
  to protect individual privacy; the absence signals "too few observations," not zero
- Test scores absent for students who dropped out — selection, not random dropout

> Always ask: *why* is this value absent? The reason shapes how you handle it.

---

## Step 4 — Anomalies and Outliers

### Box Plot + Raw Data — Spot the Outliers

```{r boxplot-jitter}
penguins |>
  ggplot(aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = species), width = 0.15, alpha = 0.4, size = 1.2) +
  scale_fill_manual(values  = c("#E69F00", "#56B4E9", "#009E73")) +
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73")) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Body Mass by Species — Spot Outliers with Box Plot + Raw Data",
       x = NULL, y = "Body Mass (g)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

> `outlier.shape = NA` suppresses duplicate outlier dots from the boxplot —
> they are already shown by `geom_jitter`. Investigate any points well beyond
> the whiskers before dropping them.

---

## Step 5 — Distribution Shape

### Histograms: The Workhorse

```{r histogram-basic}
penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram(binwidth = 2, fill = "#861F41", color = "white") +
  labs(
    title = "Distribution of Penguin Bill Length",
    x     = "Bill Length (mm)",
    y     = "Count"
  ) +
  theme_minimal(base_size = 13)
```

### Choosing `binwidth` Matters

```{r histogram-binwidth}
p_narrow <- penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram(binwidth = 0.5, fill = "#861F41", color = "white") +
  labs(title = "binwidth = 0.5 — too noisy", x = "Bill Length (mm)", y = NULL) +
  theme_minimal()

p_wide <- penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram(binwidth = 10, fill = "#861F41", color = "white") +
  labs(title = "binwidth = 10 — hides structure", x = "Bill Length (mm)", y = NULL) +
  theme_minimal()

p_narrow + p_wide
```

### Skewness and the Log Transformation

```{r log-transform}
p1 <- ggplot(state_data, aes(x = median_income)) +
  geom_histogram(bins = 20, fill = "#861F41", color = "white") +
  scale_x_continuous(labels = label_dollar(scale = 1e-3, suffix = "k")) +
  labs(title = "Raw: right-skewed", x = "Median Income", y = "Count") +
  theme_minimal(base_size = 13)

p2 <- ggplot(state_data, aes(x = log(median_income))) +
  geom_histogram(bins = 20, fill = "#E5751F", color = "white") +
  labs(title = "Log: more symmetric", x = "log(Median Income)", y = "Count") +
  theme_minimal(base_size = 13)

p1 + p2
```

### Density Plots: Smooth the Story

```{r density-species}
penguins |>
  ggplot(aes(x = bill_length_mm, fill = species)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73")) +
  labs(
    title = "Bill Length Distribution by Species",
    x     = "Bill Length (mm)",
    y     = "Density",
    fill  = "Species"
  ) +
  theme_minimal(base_size = 13)
```

### Bar Charts: Discrete Variable Distribution

```{r barplot-species}
penguins |>
  ggplot(aes(x = island, fill = species)) +
  geom_bar(width = 0.6) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73")) +
  labs(
    title = "Penguin Counts by Island and Species",
    x     = NULL,
    y     = "Count",
    fill  = "Species"
  ) +
  theme_minimal(base_size = 13)
```

---

## Step 6 — Correlations

### Scatterplot with a Trend Line

```{r scatter-delays}
set.seed(42)
flights |>
  slice_sample(n = 5000) |>
  ggplot(aes(x = dep_delay, y = arr_delay)) +
  geom_point(alpha = 0.2, color = "#861F41") +
  geom_smooth(method = "lm", color = "#E5751F", se = TRUE) +
  labs(
    title = "Departure Delay vs Arrival Delay",
    x     = "Departure Delay (min)",
    y     = "Arrival Delay (min)"
  ) +
  theme_minimal(base_size = 13)
```

### Scatterplot Matrix: `GGally::ggpairs()`

```{r ggpairs}
set.seed(42)
flights |>
  slice_sample(n = 2000) |>
  select(dep_delay, arr_delay, air_time) |>
  drop_na() |>
  ggpairs(
    lower = list(continuous = wrap("points", alpha = 0.2, color = "#861F41")),
    upper = list(continuous = wrap("cor", size = 4)),
    diag  = list(continuous = wrap("densityDiag", fill = "#E5751F", alpha = 0.5))
  ) +
  theme_minimal(base_size = 13)
```

### Correlation Heatmap: `corrplot`

```{r corrplot}
flights_num <- flights |>
  select(dep_delay, arr_delay, air_time, distance) |>
  drop_na()

corrplot(
  cor(flights_num),
  method      = "color",
  type        = "upper",
  col         = colorRampPalette(c("#861F41", "white", "#E5751F"))(200),
  tl.col      = "#2D2D2D",
  addCoef.col = "black",
  number.cex  = 0.8
)
```
