---
title: "EDA Exercise: Poverty and Income in Virginia Counties"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(skimr)
library(tidycensus)
```

## Research Question

**Q: Do counties with higher poverty rates have lower median incomes?**

We will answer this using the 6-step EDA workflow on Virginia county-level ACS data.

---

## Load the Data

Run `get_acs()` once (requires a free Census API key). If you don't have one yet,
register at <https://api.census.gov/data/key_signup.html> and run:
`tidycensus::census_api_key("YOUR_KEY", install = TRUE)`

```{r load-data}
va_counties <- get_acs(
  geography = "county",
  state     = "VA",
  variables = c(
    median_income = "B19013_001",   # median household income
    pov_count     = "B17001_002",   # people below poverty line
    pov_total     = "B17001_001",   # total for whom poverty status determined
    pct_college   = "B15003_022"    # bachelor's degree holders (25+)
  ),
  year   = 2022,
  output = "wide"
) |>
  mutate(
    pov_rate    = pov_countE / pov_totalE,
    county_name = str_remove(NAME, ", Virginia")
  )

glimpse(va_counties)
```

---

## Step 1 — Summary Statistics

Use `skim()` to get a full picture of the dataset before anything else.

```{r step1-skim}
va_counties |>
  select(county_name, median_incomeE, pov_rate, pct_collegeE) |>
  skim()
```

**Questions to answer:**

1. How many counties are in this dataset? How many variables?
2. Are there any variables with missing values (`n_missing > 0`)? Which ones?
3. What is the range of `median_incomeE`? Does the minimum or maximum look suspicious?
4. What is the range of `pov_rate`? Does it stay between 0 and 1?

*Your answers here:*


---

## Step 2 — Spot Problems in the Numbers

Check for missingness and impossible values.

```{r step2-missing}
# Which variables have missing values?
va_counties |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0)
```

```{r step2-range}
# Check value ranges
va_counties |>
  summarise(
    income_min = min(median_incomeE, na.rm = TRUE),
    income_max = max(median_incomeE, na.rm = TRUE),
    pov_min    = min(pov_rate, na.rm = TRUE),
    pov_max    = max(pov_rate, na.rm = TRUE)
  )
```

**Questions to answer:**

5. Are any poverty rates outside the range [0, 1]? If so, which counties?
6. Is there any missingness that concerns you? What might explain it?

*Your answers here:*


---

## Step 3 — Hidden Missing Values

Check for sentinel codes and implicitly missing data.

```{r step3-sentinel}
# Are any income values coded as 0 or negative (common data error)?
va_counties |>
  filter(median_incomeE <= 0 | is.na(median_incomeE)) |>
  select(county_name, median_incomeE)
```

```{r step3-implicit}
# Do all 133 Virginia counties appear?
# (Expected: 133 counties + some independent cities)
n_distinct(va_counties$GEOID)
```

**Questions to answer:**

7. Are there any zero or negative income values?
8. Are any counties missing entirely from the dataset? What might explain that?

*Your answers here:*


---

## Step 4 — Anomalies and Outliers

Use a box plot to identify outlier counties.

```{r step4-boxplot}
va_counties |>
  ggplot(aes(y = median_incomeE)) +
  geom_boxplot(fill = "#861F41", alpha = 0.6, outlier.color = "#E5751F", outlier.size = 2) +
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-3, suffix = "k")) +
  labs(
    title = "Distribution of Median Household Income — Virginia Counties",
    y     = "Median Household Income"
  ) +
  theme_minimal(base_size = 13)
```

```{r step4-identify}
# Which counties are outliers (above 75th percentile + 1.5 × IQR)?
q75  <- quantile(va_counties$median_incomeE, 0.75, na.rm = TRUE)
q25  <- quantile(va_counties$median_incomeE, 0.25, na.rm = TRUE)
iqr  <- q75 - q25

va_counties |>
  filter(median_incomeE > q75 + 1.5 * iqr) |>
  select(county_name, median_incomeE) |>
  arrange(desc(median_incomeE))
```

**Questions to answer:**

9. Which counties appear as high-income outliers? Are they data errors or real extreme cases?
10. Should you drop them or keep them? Why?

*Your answers here:*


---

## Step 5 — Distribution Shape

Examine the shape of your key variables.

```{r step5-hist-income}
va_counties |>
  ggplot(aes(x = median_incomeE)) +
  geom_histogram(bins = 25, fill = "#861F41", color = "white") +
  scale_x_continuous(labels = scales::label_dollar(scale = 1e-3, suffix = "k")) +
  labs(
    title = "Distribution of Median Household Income",
    x     = "Median Income",
    y     = "Count"
  ) +
  theme_minimal(base_size = 13)
```

```{r step5-hist-pov}
va_counties |>
  ggplot(aes(x = pov_rate)) +
  geom_histogram(bins = 25, fill = "#E5751F", color = "white") +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(
    title = "Distribution of Poverty Rate",
    x     = "Poverty Rate",
    y     = "Count"
  ) +
  theme_minimal(base_size = 13)
```

```{r step5-log}
# Try log transformation on income — does it look more symmetric?
va_counties |>
  ggplot(aes(x = log(median_incomeE))) +
  geom_histogram(bins = 25, fill = "#861F41", color = "white") +
  labs(
    title = "log(Median Household Income) — More Symmetric?",
    x     = "log(Median Income)",
    y     = "Count"
  ) +
  theme_minimal(base_size = 13)
```

**Questions to answer:**

11. Is `median_incomeE` right-skewed, left-skewed, or roughly symmetric?
12. Is `pov_rate` right-skewed, left-skewed, or roughly symmetric?
13. Does the log transformation of income look more symmetric? Would you use it in a regression?

*Your answers here:*


---

## Step 6 — Correlations

**Answer the research question: Do counties with higher poverty rates have lower median incomes?**

```{r step6-scatter}
va_counties |>
  ggplot(aes(x = pov_rate, y = median_incomeE)) +
  geom_point(color = "#861F41", alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", color = "#E5751F", se = TRUE) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-3, suffix = "k")) +
  labs(
    title   = "Poverty Rate vs Median Income — Virginia Counties, 2022",
    x       = "Poverty Rate",
    y       = "Median Household Income",
    caption = "Source: ACS 5-Year Estimates 2022"
  ) +
  theme_minimal(base_size = 13)
```

```{r step6-correlation}
# Compute the correlation coefficient
cor(va_counties$pov_rate, va_counties$median_incomeE, use = "complete.obs")
```

```{r step6-label-outliers}
# Which counties stand out? Label them.
library(ggrepel)

high_pov  <- va_counties |> slice_max(pov_rate, n = 5)
high_inc  <- va_counties |> slice_max(median_incomeE, n = 5)
labeled   <- bind_rows(high_pov, high_inc) |> distinct(GEOID, .keep_all = TRUE)

va_counties |>
  ggplot(aes(x = pov_rate, y = median_incomeE)) +
  geom_point(color = "#861F41", alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", color = "#E5751F", se = TRUE) +
  geom_label_repel(data = labeled, aes(label = county_name),
                   size = 2.8, box.padding = 0.4) +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_dollar(scale = 1e-3, suffix = "k")) +
  labs(
    title   = "Poverty Rate vs Median Income — Virginia Counties, 2022",
    x       = "Poverty Rate",
    y       = "Median Household Income",
    caption = "Source: ACS 5-Year Estimates 2022"
  ) +
  theme_minimal(base_size = 13)
```

**Questions to answer:**

14. What is the direction and magnitude of the correlation (r = ?)?
15. In one sentence, describe what the scatterplot shows.
16. Are there any counties that don't fit the general pattern? Which ones, and why might they be exceptions?
17. Is this a causal relationship or just an association? What other variables might explain both poverty rate and income?

*Your answers here:*


---

## Bonus: Apply to Your Project Data

Replace `va_counties` with your own project data and repeat the 6 steps:

```{r bonus-template, eval=FALSE}
# Load your project data
my_data <- read_csv("path/to/your/data.csv")

# Step 1: skim
my_data |> skim()

# Step 2: Check missingness and ranges
my_data |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything()) |>
  filter(value > 0)

# Step 3: Sentinel codes?
# my_data |> mutate(col = na_if(col, 9999))

# Step 4: Boxplot of your outcome variable
# my_data |> ggplot(aes(y = YOUR_OUTCOME)) + geom_boxplot()

# Step 5: Histogram of key variables
# my_data |> ggplot(aes(x = YOUR_VARIABLE)) + geom_histogram()

# Step 6: Scatterplot of your main relationship
# my_data |> ggplot(aes(x = YOUR_PREDICTOR, y = YOUR_OUTCOME)) +
#   geom_point() + geom_smooth()
```
